Having established the soundness of our translation, it is time 
we see in this section how we can use this sound translation to verify a program. 
The following theorem is then true:

\begin{theorem}[Soundness]\label{thm:prover-soundness}
Assume that $e$ and $\Ct$ contain only function symbols from $P$ and no free term variables.
Let $\Th_{all} = \Th\;\land\;\ptrans{}{P}$. 
If $\Th_{all} \land \neg\ctrans{\Sigma}{P}{e \in \Ct}$ is unsatisfiable 
then $\langle D_\infty,{\cal I}\rangle \models \ctrans{\Sigma}{P}{e \in \Ct}$ and 
consequently $\dbrace{e} \in \dbrace{\Ct}$.
\end{theorem}
\begin{proof}
If there is no model for this formula then its negation must be valid (true in all models), that 
is $ \neg \Th_{all} \lor \ctrans{\Sigma}{P}{e \in \Ct}$ is valid. By completeness
of first-order logic $\Th_{all} |- \ctrans{\Sigma}{P}{e \in \Ct}$. This means 
that all models of $\Th_{all}$ validate $\ctrans{\Sigma}{P}{f \in \Ct}$. In particular, 
for the denotational model we have that $\langle D_\infty,{\cal I}\rangle \models \Th_{all}$ 
and hence $\langle D_\infty,{\cal I} \rangle \models \ctrans{\Sigma}{P}{e \in \Ct}$. 
Theorem~\ref{thm:den-contr-satisfaction} finishes the proof.
\end{proof}

Hence, to verify a program $e$ satisfies a contract $\Ct$ we need to do the following:
\begin{itemize}
  \item Generate formulae for the theory $\Th\;\land\;\ptrans{}{P}$
  \item Generate the negation of a contract translation: $\neg\ctrans{\Sigma}{P}{e \in \Ct}$
  \item Ask a SAT solver for a model for the conjunction of the above formulae
\end{itemize}

\paragraph{Incremental verification}

Theorem~\ref{thm:prover-soundness} gives us a way to check that an expression satisfies a 
contract. Assume that we are given a program $P$ with a function $f \in dom(P)$, for which 
we have already shown that $\langle D_\infty,{\cal I}\rangle \models \ctrans{\Sigma}{P}{f \in \Ct_f}$.
Suppose next that we are presented with a ``next'' goal, to prove that 
$\langle D_\infty,{\cal I}\rangle \models \ctrans{\Sigma}{P}{h \in \Ct_h}$. 
We may consider the following three variations of how to do this:

\begin{itemize}
  \item Ask for the unsatisfiability of: 
    \[  \Th\; \land \; \ptrans{\Sigma}{P} \; \land \;\neg\ctrans{\Sigma}{P}{h \in \Ct_h} \]
        The soundness of this query follows from Theorem~\ref{thm:prover-soundness} above.

  \item Ask for the unsatisfiability of:
    \[  \Th \; \land\; \ptrans{\Sigma}{P} \; \land \; \ctrans{\Sigma}{P}{f \in \Ct_f} \land \neg \ctrans{\Sigma}{P}{h \in \Ct_h} \]
        This query adds the {\em already proven} contract for $f$ to the theory. If this formula
        is unsatisfiable, then its negation is valid, and we know that the denotational model is 
        a model of the theory {\em and} of $\ctrans{\Sigma}{P}{f \in \Ct_f}$ and hence it must also
        be a model of $\ctrans{\Sigma}{P}{h \in \Ct_h}$.
  \item Ask for the unsatisfiability of:
    \[  \Th \; \land \; 
        \ptrans{\Sigma}{P \setminus f} \; \land \; 
        \ctrans{\Sigma}{P}{f \in \Ct_f} \; \land \;
        \neg\ctrans{\Sigma}{P}{h \in \Ct_h} \]
        This query removes the axioms associated with the \emph{definition} of $f$, leaving
        only its \emph{contract} available.  This makes the proof of $h$'s contract
        insensitive to changes in $f$'s immplementation.
        Via a similar reasoning as before, such an invocation 
        is sound as well.
\end{itemize}


%% Our final goal is going to show that a program does not crash, that
%% is the final contract will be of the form $e \in \Ct$ where $\Ct$ is
%% going to be some {\em base contract}. Note that by base contract adequacy
%% (Lemma~\ref{lem:base-contract-adequacy}) if we manage to show a base contract 
%% denotationally, then the contract holds in operational terms.

%% \clearpage
